---
title: "P8131_hw_2"
author: "Renjie Wei rw2844"
date: "2/18/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
#options(knitr.table.format = "latex")
```

## Problem 1
Fit the model $g(\text{P}(\text{dying})) = \alpha +\beta X$, with logit, probit, and complementary log-log links.

(a) Fill out the table and give comments.

```{r problem_1_a, echo=FALSE}
data1 <- data.frame(
  dose = c(0,1,2,3,4),
  nd = c(2,8,15,23,27),
  na = 30-c(2,8,15,23,27)
)

logit.fit <- glm(cbind(nd,na) ~ dose, family = binomial(link = "logit"), data = data1)
probit.fit <- glm(cbind(nd,na) ~ dose, family = binomial(link = "probit"), data = data1)
cloglog.fit <- glm(cbind(nd,na) ~ dose, family = binomial(link = "cloglog"), data = data1)

library(tidyverse)
summary_tab_1 <- 
  data.frame(
    model = c("Logit", "Probit", "c-log-log"),
    est_beta = c(logit.fit$coefficients[[2]],probit.fit$coefficients[[2]],cloglog.fit$coefficients[[2]]),
    CI_beta_L = round(c(
      logit.fit$coefficients[[2]]-(1.96*(summary(logit.fit)$coefficients[2,2])),
      probit.fit$coefficients[[2]]-(1.96*(summary(probit.fit)$coefficients[2,2])),
      cloglog.fit$coefficients[[2]]-(1.96*(summary(cloglog.fit)$coefficients[2,2]))
    ),3),
    CI_beta_U = round(c(
      logit.fit$coefficients[[2]]+(1.96*(summary(logit.fit)$coefficients[2,2])),
      probit.fit$coefficients[[2]]+(1.96*(summary(probit.fit)$coefficients[2,2])),
      cloglog.fit$coefficients[[2]]+(1.96*(summary(cloglog.fit)$coefficients[2,2]))
    ),3),
    Deviance = c(
      logit.fit$deviance,
      probit.fit$deviance,
      cloglog.fit$deviance
    ),
    p_ = c(
      predict(logit.fit,newdata = data.frame(dose = 0.01), type = "response"),
      predict(probit.fit,newdata = data.frame(dose = 0.01), type = "response"),
      predict(cloglog.fit,newdata = data.frame(dose = 0.01), type = "response")
    )
    
  )
summary_tab_1$CI = c(
      paste(summary_tab_1$CI_beta_L[1],summary_tab_1$CI_beta_U[1],sep = ","),
      paste(summary_tab_1$CI_beta_L[2],summary_tab_1$CI_beta_U[2],sep = ","),
      paste(summary_tab_1$CI_beta_L[3],summary_tab_1$CI_beta_U[3],sep = ",")
    )

summary_tab_1 %>%
  select(model,est_beta,CI,Deviance,p_) %>% 
  knitr::kable(col.names = c("Model", "Estimate of $\\beta$", expression("95% CI of $\\beta$"),"Deviance","p(dying|x=0.01)"), digits = 3)
```

Comments:

- All the estimated $\beta$s are greater than 0, which means the increase in dose may increase the probability of dying. 
- 0 is not included in all the 3 95% CIs, that means we have 95% confidence to conclude that the dose level is significantly associated with the probability of dying.
- All the deviances follow the $\large{\chi^2}(3)$, the deviance of the
probit link model is the smallest, which means it is the best fitted model among all the three models.
- About the probability of dying conditioned on 0.01 dose, the logit link model gives us a 0.0901,the probit link model gives a 0.0853 and the c-log-log link model gives a 0.1282.


(b) Suppose that the dose level is in natural logarithm scale, estimate LD50 with 90%
confidence interval based on the three models.

Since we got the following link functions:

Logit link function:

$$g_1(\pi) = log(\frac{\pi}{1-\pi}) = \beta_0+\beta_1x$$
Probit link function:
$$g_2(\pi) = \Phi^{-1}(\pi)= \beta_0+\beta_1x$$
C-log-log link function:
$$g_3(\pi) = \log(-\log(1-\pi))= \beta_0+\beta_1x$$
So we can get the point estimates of $LD50$ by solving the following equation:

$$g(0.5)=\beta_0+\beta_1x$$
Logit estimate:
$$\hat x = -\frac{\hat{\beta_0}}{\hat\beta_1} = f({\bf {\hat{\beta}}})$$
Probit estimate:
$$\hat x = \frac{\Phi^{-1}(0.5)-\hat{\beta_0}}{\hat\beta_1}= f({\bf {\hat{\beta}}})$$
C-log-log estimate:
$$\hat x = \frac{\log(-\log(0.5))-\hat{\beta_0}}{\hat\beta_1}= f({\bf {\hat{\beta}}})$$
And we can get the asymptotic variance of $\hat x$:
$$\text{var}(\hat x) = \text{var}(f({\bf {\hat{\beta}}})) = (\frac{\partial f({\bf{\hat{\beta}}})}{\partial \beta_0})^2\text{var}(\hat\beta_0)+(\frac{\partial f({\bf{\hat{\beta}}})}{\partial \beta_1})^2\text{var}(\hat\beta_1)+2(\frac{\partial f({\bf{\hat{\beta}}})}{\partial \beta_0})(\frac{\partial f({\bf{\hat{\beta}}})}{\partial \beta_1})\text{cov}(\hat\beta_0,\hat\beta_1)$$
Then the asymptotic CI of $LD50$ is:
$$LD50 \in [e^{\hat x-\text{z}_{\alpha/2}\sqrt{\text{var}(\hat x)}},e^{\hat x+\text{z}_{\alpha/2}\sqrt{\text{var}(\hat x)}}]$$
I write a function to calculate the $\text{var}(\hat x)$, which is `se_of_h_beta`:

```{r problem-1-b}
# a function to calculate se for any h(beta)
se_of_h_beta <- function(fit_object, h_expr){
  beta_0 = fit_object$coefficients[["(Intercept)"]]
  beta_1 = fit_object$coefficients[[2]]
  h_beta = eval(h_expr)
  
  I_beta = vcov(fit_object)
  #inv_I_beta = solve(I_beta) already inversed!
  partial_d_beta0 = eval(D(h_expr,"beta_0"))
  partial_d_beta1 = eval(D(h_expr,"beta_1"))
  
  partial_d_mtx = matrix(c(partial_d_beta0,partial_d_beta1),2,1)
  se = sqrt(t(partial_d_mtx) %*% I_beta %*% partial_d_mtx)
  return(c(h_beta,se))
}

LD50_logit = expression(-beta_0/beta_1)
e_logit = se_of_h_beta(logit.fit, LD50_logit)


LD50_probit = expression(-beta_0/beta_1)
e_probit = se_of_h_beta(probit.fit, LD50_probit)

g_pi_cloglog = log(-log(0.5))
LD50_cloglog = expression((g_pi_cloglog-beta_0)/beta_1)
e_cloglog = se_of_h_beta(cloglog.fit, LD50_cloglog)

summary_tab_1_b <- 
  data.frame(
    Model = c("Logit", "Probit", "c-log-log"),
    LD50 = c(exp(e_logit[1]), exp(e_probit[1]), exp(e_cloglog[1])),
    CI_L = exp(c(e_logit[1], e_probit[1], e_cloglog[1]) - qnorm(.95)*c(e_logit[2], e_probit[2], e_cloglog[2])),
    CI_U = exp(c(e_logit[1], e_probit[1], e_cloglog[1]) + qnorm(.95)*c(e_logit[2], e_probit[2], e_cloglog[2]))
  )


```

The following table shows the results of estimated LD50.  
```{r summary_LD50, echo=FALSE}
summary_tab_1_b %>% 
  knitr::kable(col.names = c("Model","Estimate of LD50","90% CI lower", "90% CI upper"),digit=3)
```



## Problem 2
Please analyze the data using a logistic regression and answer the following questions.

(a) How does the model fit the data?

Firstly, I made a dataframe for the data.
```{r problem-2-a-1, echo=FALSE}
mph_enroll <- 
  data.frame(
    amount = seq(10, 90, by = 5),
    offers = c(4, 6, 10, 12, 39, 36, 22, 14, 10, 12, 8, 9, 3, 1, 5, 2, 1),
    enrolls = c(0, 2, 4, 2, 12, 14, 10, 7, 5, 5, 3, 5, 2, 0, 4, 2, 1)
  )
```

We assume the response of those who received offers $Y_i$ in each group follow the same Bernoulli distribution $Y_i \sim Bin(1,\pi)$, then in each group $j$ with size $m_j$, $Y = \sum\limits_{i=1}^{m_j}Y_i$ has a Binomial distribution, that is $Y\sim Bin(m_j,\pi)$.

So we fit a logistic regression based on the above assumptions.
```{r mph_fit}
mph.fit <-
  glm(cbind(enrolls, offers-enrolls) ~ amount, family = binomial(link = "logit"), data = mph_enroll)
```

And I do Hosmer–Lemeshow test for the goodness-of-fit since the data is sparse.
```{r hoslem_test}
library(ResourceSelection)
hoslem.test(mph.fit$y, fitted(mph.fit), g=10)
```
The result shows that the Hosmer-Lemeshow stastic $\chi^2_{HL}$ is `X-squared = 1.6111` with `df = 8` and `p-value = 0.9907`. Since `p-value` is greater than 0.05, we fail to reject the null hypothesis and conclude that there is no evidence of that the model is lack of fit.


(b) How do you interpret the relationship between the scholarship amount and enrollment
rate? What is 95% CI?

```{r summary_mph_fit, echo=FALSE}
summary(mph.fit)
e_beta = mph.fit$coefficients[[2]]
OR = exp(e_beta)
se_beta = summary(mph.fit)$coefficients["amount","Std. Error"]
beta_CI_L = e_beta - qnorm(.975)*se_beta
beta_CI_U = e_beta + qnorm(.975)*se_beta
OR_CI_L = exp(beta_CI_L)
OR_CI_U = exp(beta_CI_U)
```

The estimate coefficient $\beta_1$ is `r round(e_beta,3)`, which is equal to the $\log(OR_{\text{amout}})$. So the $OR_{\text{amout}}$ is `r round(exp(e_beta),3)`, which means the odds ratio of enrollment increases 3.14% per 1000 dollars increase in the scholarship. The 95% CI of OR is (`round(OR_CI_L,3)`,`round(OR_CI_U,3)`)


(c) How much scholarship should we provide to get 40% yield rate (the percentage of
admitted students who enroll?) What is the 95% CI?

Since we use logit link function, we can get the estimate of scholarship (to get 40% yield rate) by solving the following equation:
$$\log(\frac{\pi}{1-\pi}) = \log(0.4/0.6) = \hat\beta_0+\hat\beta_1\times\widehat{\text{scholarship}}$$
Then we can get the estimate scholarship and the asymptotic variance and CI with the same method in problem 1.
  
```{r mph_estimate, echo=FALSE}
mph_expr = expression((log(2/3)-beta_0)/beta_1)
e_mph = se_of_h_beta(mph.fit, mph_expr)
e_mph_CI_L = e_mph[1] - qnorm(.975)*e_mph[2]
e_mph_CI_U = e_mph[1] + qnorm(.975)*e_mph[2]
```

We get the estimate of scholarship $\widehat {\text{scholarship}}$ = `r round(e_mph[1],3)`, which means we should provide `r round(e_mph[1],3)` thousand dollars of scholarship to get 40% yield rate. The 95% CI is (`r round(e_mph_CI_L,3)`,`r round(e_mph_CI_U,3)`)