---
title: "P8131 HW8"
author: "Renjie Wei rw2844"
output: pdf_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = F, 
                      message = F, 
                      warning = F, 
                      comment = "")
library(tidyverse)
library(readxl)
library(gee)
library(lme4)
library(nlme)
```

The data (HEALTH.xlsx) are from a randomized, controlled trial among women of childbearing age to evaluate the effects of an educational intervention. One response variable of interest is the participants self-rating of health status as either good or poor. The researchers would like to assess the effect of the intervention on self-rated health across the follow-up period, as well as whether these effects are influenced by the mothers age. There are n = 80 women enrolled in this trial. These data
were measured at 4 points in time: randomization, 3 months, 6 months, and 12 months post-randomization.


```{r}
health.df = read_excel("./HW8-HEALTH.xlsx") %>%
  rename(TRT = TXT) %>%
  mutate(TRT = as.factor(TRT),
         HEALTH = as.numeric(HEALTH == "Good"),
         TIME = as.integer(TIME))
```

\newpage

## Problem a

Evaluate the bivariate, cross-sectional relationship between randomized group assignment and participants health self-rating at the time of randomization. Interpret and discuss these findings.

### Answers

```{r}
randomized.df = health.df %>% filter(TIME == 1)
randomized.fit = glm(HEALTH ~ TRT,
                  data = randomized.df,
                  family = binomial(link = "logit"))
randomized.fit %>% broom::tidy(conf.int = T) %>% knitr::kable(digits = 3, caption = "Logistic Model Coefficients")
#randomized.fit %>% broom::glance() %>% knitr::kable(caption = "Logistic Model Goodness-of-Fit statistics")
```

The odds ratio of self-reporting "good" health for status intervention group vs. control group at randomization (baseline) is `r round(exp(randomized.fit$coefficients[2]), 3)`. However, the p value for the coefficient is 0.486 > 0.05; therefore, we are 95% confident to conclude that there is not enough evidence to support association between treatment group assignment and health status at randomization. 


\newpage

## Problem b 

Perform a longitudinal data analysis across all study follow-up visits (but not at randomization) to describe the relationship of the participants self-ratings as a function of the effects of health self-rating at the baseline, treatment group, month post randomization, and age group as predictors. Fit a GEE with unstructured correlation structure. Interpret your results.

### Answers

I transformed the `TIME` to its original scale, that is $0,3,6,12$.

The GEE model we are considering is
$$\log\{\frac{\Pr(Y_{ij} = 1|X_{ij})}{\Pr(Y_{ij} = 0|X_{ij})}\} =\log(\mu_{ij})= \beta_0+\beta_1\times\text{BASELINEGood}_i+\beta_2\times\text{Intervention}_i+\beta_3\times\text{TIME}_i+\beta_4\times\text{AGE:25-34}_i+\beta_5\times\text{AGE:35+}_i$$
since the study design is unbalanced plus the response is binary
$$\text{Var}{(Y_{ij} |X_{ij})} = \phi v(\mu_{ij}) = \mu_{ij}(1-\mu_{ij})$$
finally, the within-subject association among the vector of repeated responses is assumed to have an unstructured pairwise log odds ratio pattern,

$$\log \text{OR}(Y_{ij},Y_{ik}|X_{ij},X_{ik}) = \alpha_{jk}$$

```{r, include = F}
# remove the participants with randomization only, which is 108 and 115
health.lda = health.df %>%
    filter(!ID %in% names(which(table(health.df$ID) == 1)))
health.baseline = subset(health.lda, TIME == "1")
health.lda$BASELINE = rep(health.baseline$HEALTH, as.numeric(table(health.lda$ID)))
health.lda$TIME = health.lda$TIME - 1
health.lda = subset(health.lda, TIME > 0)
health.lda$TIME = case_when(health.lda$TIME == 1 ~ 3,
                            health.lda$TIME == 2 ~ 6,
                            health.lda$TIME == 3 ~ 12)
   
gee.fit = gee(HEALTH ~ BASELINE + TRT + TIME + AGEGROUP,
              data = health.lda,
              id = ID,
              family = "binomial",
              corstr = "unstructured",
              scale.fix = TRUE,
              scale.value = 1,
              
              )

```
```{r}
gee.sum = summary(gee.fit)
gee.sum$coefficients %>% knitr::kable(digits = 3, caption = "Parameter estimates and standard errors from GEE model")
data.frame(gee.sum$working.correlation) %>% mutate(obs = c("2", "3","4")) %>% select(obs,everything()) %>% knitr::kable(digits = 3, caption = "Working correlation estimates from GEE model", col.names = c("obs", "2", "3","4"))
```

**Interpretations:**

* The odds ratio of participants self-rating "good" vs. "poor" is `r round(exp(gee.fit$coefficients[2]), 3)`, between participants self-rating "good" or "poor" at baseline, if take average among all measurements and all subjects within the same subgroup. 

* The odds ratio of participants self-rating "good" vs. "poor" is `r round(exp(gee.fit$coefficients[3]), 3)`, between participants in "intervention" or "control" treatment group, if take average among all measurements and all subjects within the same subgroup. 

* The odds ratio of participants self-rating "good" vs. "poor" is `r round(exp(gee.fit$coefficients[4]), 3)`, for per month after randomization change, if take average among all measurements and all subjects within the same subgroup. 

* The odds ratio of participants self-rating "good" vs. "poor" is `r round(exp(gee.fit$coefficients[5]), 3)`, between 25-34 age group vs. 15-24 age group, if take average among all measurements and all subjects within the same subgroup. 

* The odds ratio of participants self-rating "good" vs. "poor" is `r round(exp(gee.fit$coefficients[6]), 3)`, between 35+ age group vs. 15-24 age group, if take average among all measurements and all subjects within the same subgroup. 

\newpage

## Problem c

Fit a generalized linear mixed effects model with subject-specific random intercepts. Interpret your estimates. How are the interpretations different from the GEE model?

### Answers

The GLMM model we are considering is
$$\log\{\frac{\Pr(Y_{ij} = 1|b_{i})}{\Pr(Y_{ij} = 0|b_{i})}\} = \beta_0+\beta_1\times\text{BASELINEGood}_{ij}+\beta_2\times\text{Intervention}_{ij}+\beta_3\times\text{TIME}_{ij}+\beta_4\times\text{AGE:25-34}_{ij}+\beta_5\times\text{AGE:35+}_{ij} + b_i$$

```{r}
glmm.fit = glmer(HEALTH ~ BASELINE + TRT + TIME + AGEGROUP + (1 | ID),
             data = health.lda,
             family = binomial)
glmm.sum = summary(glmm.fit)
glmm.sum$coefficients %>% knitr::kable(digits = 3, caption = "Estimated fixed effect for random intercept model")
as.data.frame(as.matrix(glmm.sum$vcov)) %>% knitr::kable(digits = 3, caption = "Covariance matrix for the estimates of fixed effects")
as.data.frame(glmm.sum$varcor) %>% select(-var2) %>% knitr::kable(digits = 3, caption = "Estimated random intercept", col.names = c("Group","Variable","vcov","sd"))
```

**Interpretation: **

* For the same subject, the odds ratio of participants self-rating "good" vs. "poor" is `r round(exp(coef(glmm.fit)[[1]][1,4]), 3)`, for per month after randomization change, while holding other covariates fixed.

* For the same subject, The odds ratio of participants self-rating "good" vs. "poor" is `r round(exp(coef(glmm.fit)[[1]][1,2]), 3)`, for participants self-rating "good" vs, "poor" at baseline, while holding other covariates fixed.

* For the same subject, The odds ratio of participants self-rating "good" vs. "poor" is `r round(exp(coef(glmm.fit)[[1]][1,3]), 3)`, for participants in "intervention" vs. "control" treatment group, while holding other covariates fixed.

* For the same subject, The odds ratio of participants self-rating "good" vs. "poor" is `r round(exp(coef(glmm.fit)[[1]][1,5]), 3)`, for participants in 25-34 age group vs. 15-24 age group, while holding other covariates fixed.

* For the same subject, The odds ratio of participants self-rating "good" vs. "poor" is `r round(exp(coef(glmm.fit)[[1]][1,6]), 3)`, for participants in 35+ age group vs. 15-24 age group, while holding other covariates fixed.

**Difference between the GEE and GLMM parameter interpretations: **

* GEE is a marginal model focus on inferences about the study population.

* GLMM focus on inference about individuals.

* The parameters in GLMM measure the change in expected value of response while holding other covariates and *random effect* constant. On the contrary, the parameters in GEE measures the change in expected value of response, for *sub-populations* defined by fixed and known covariate values.

## Appendix

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```

