---
title: "P8131 HW9"
author: "Renjie Wei rw2844"
output: 
  pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F)
```


\newpage

## Problem 1

Determine the survival and density functions for a continuous survival time variable
with hazard function

$$h(x) = \frac{2x}{(1+x^2)}$$

### Answers

The cumulative harzard funciton is

$$
\begin{aligned}
H(t) &= \int\limits_0^{t}h(x)dx \\ 
& =\int\limits_0^{t}\frac{2x}{(1+x^2)} \\
& = \log(1+t^2)\\
\end{aligned}
$$
The survival function is 

$$
\begin{aligned}
S(t) &= \exp(-H(t)) \\ 
& = \exp(-\log(1+t^2)) \\ 
& = \frac{1}{1+t^2}
\end{aligned}
$$

$$
\begin{aligned}
F(t) &= 1-S(t) \\ 
& = \frac{t^2}{1+t^2}
\end{aligned}
$$

Hence, the density function is

$$
\begin{aligned}
f(t) &= F'(t) \\ 
& = (1-\frac{1}{1+t^2})' \\ 
& = \frac{2t}{(1+t^2)^2}
\end{aligned}
$$
\newpage

## Problem 2

### a

Find the Kaplan-Meier estimate of the survival function

### Answers

The Kaplan-Meier estimator of survival function is

$$\hat S(t) = \prod\limits_{i =1}^k(1-\hat\lambda_i) = \prod\limits_{i =1}^k(1-\frac{d_i}{n_i})$$

### b

Find the Nelson-Aalen estimate of the cumulative hazard function

### Answers

The Nelson-Aalen estimate of the cumulative hazard function is 
$$\tilde H(t) = \sum \limits_{t_i\leq t}\frac{d_i}{n_i}$$


### c

Find the Fleming-Harrington estimate of the survival function

### Answers

The Fleming-Harrington estimate of the survival function $S(t)$ is 
$$\exp(-\tilde H(t))$$

The three estimator above can be shown in the table

```{r q2_tab}
q2_tab <- data.frame(
    ti = c(1,2,4,5,6,7,8,9,10),
    ni = c(10,9,7,6,5,4,3,2,1),
    di = c(1,2,0,0,1,0,0,0,0),
    ci = c(0,0,1,1,0,1,1,1,1)
)

q2_tab$lambda <- ifelse(q2_tab$ni == 0, 0, q2_tab$di/q2_tab$ni)

for (i in 1:nrow(q2_tab)){
    q2_tab$St[i] = prod(1 - q2_tab$lambda[1:i])
}

for (i in 1:nrow(q2_tab)){
    q2_tab$Ht[i] = sum(q2_tab$lambda[1:i])
}

q2_tab$expHt <- exp(-q2_tab$Ht)
```
```{r q2_tab_kable}
library(tidyverse)
names(q2_tab) = c(
                 "$t_i$",
                 "$n_i$",
                 "$d_i$",
                 "$c_i$",
                 "$\\hat{\\lambda_i}$",
                 "$\\hat{S(t)}$",
                 "$\\tilde{H}(t)$",
                 "$\\exp(-\\tilde{H}(t))$"
             )
q2_tab %>% knitr::kable(
             digits = 3,
             escape = FALSE,
             row.names = FALSE,
             caption = "Table of the survival data 1,2,2,4+,5+,6,7+,8+,9+,10+"
             ) %>% kableExtra::footnote(general = 
                                       c(
                             "$\\\\hat{S(t)}$: Kaplan-Meier estimator of survival function ",
                             "$\\\\tilde{H(t)}$: Nelson-Aalen estimate of the cumulative hazard function",
                             "$\\\\exp(-\\\\tilde{H(t)})$: Fleming-Harrington estimate of the survival function"
                         ),
                         escape = F)
```

\newpage

## Problem 3

Use the tongue data in the R package KMsurv. For each tumor type (aneuploidy and diploid), plot the Kaplan-Meier curve of survival function and its pointwise 95% confidence intervals (using the log transformation). What are the estimated 1-year survival rate and 95% CI?

### Answers

```{r tongue}
require(KMsurv)
require(survival)
data(tongue)
aneuploid = tongue %>% filter(type == 1) %>% mutate(survt = Surv(time, delta, type = "right"))
diploid = tongue %>% filter(type == 2) %>% mutate(survt = Surv(time, delta, type = "right"))
```

The Kaplan-Meier curve of aneuploid tumor survival function and its pointwise 95% confidence intervals
```{r km_aneuploid}
KM_aneuploid = survfit(Surv(time, delta) ~ 1, data = aneuploid, conf.type = "log")
plot(KM_aneuploid, conf.int = TRUE, mark.time = TRUE, xlab = "Weeks", ylab = "Survival Probability", main = "K-M curve of aneuploid tumor", cex.lab = 1.5, cex.main = 1.5)
```

The Kaplan-Meier curve of diploid tumor survival function and its pointwise 95% confidence intervals
```{r km_diploid}
KM_diploid = survfit(Surv(time, delta) ~ 1, data = diploid, conf.type = "log")
plot(KM_diploid, conf.int = TRUE, mark.time = TRUE, xlab = "Weeks", ylab = "Survival Probability", main = "K-M curve of diploid tumor", cex.lab = 1.5, cex.main = 1.5)
```

The table below summarizes the estimated 1-year survival rate and $95\%$ CI.

```{r one_year_surv_prob}
sum_an <- summary(KM_aneuploid, time = c(52))
sum_di <- summary(KM_diploid, time = c(52))
one_year_surv <- data.frame(
    tumor = c("aneuploid tumor", "diploid tumor"),
    rate = c(sum_an$surv, sum_di$surv),
    lower = c(sum_an$lower, sum_di$lower),
    upper = c(sum_an$upper, sum_di$upper)
)

one_year_surv %>% knitr::kable(digits = 3,
                              caption = "Estimated 1-year survival rate and $95\\%$ CI of different tumor types",
                              escape = F,
                              row.names = F,
                              col.names = c(
                                  "Tumor Type",
                                  "1-year survival rate",
                                  "lower $95\\%$ CI",
                                  "upper $95\\%$ CI"
                              ),
                              format = "latex"
                              )
```




## Appendix

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```

